---
title: "Storm Data Analysis"
author: "Levshinovskiy Mikhail"
date: "1/25/2021"
output: html_document
---

## Data Processing

Read dataset from archive, check if dimensions of data frame are correct. 
(Expected value: 902297 x 37)
```{r read_storm_data, cache=TRUE}
storm_data <- read.csv("./repdata_data_StormData.csv.bz2")
print(dim(storm_data))
names(storm_data)
```

## Results
  
### Question 1
  *Quesion* : Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
  
To answer this question we will sum number of injures and fatal cases for every event
type in dataset. Below numbers and *barplot* for 5 the most harmful events are presented.

```{r count_most_harmful_type}
fatal_cases_by_type <- sort(tapply(storm_data$FATALITIES, storm_data$EVTYPE, sum),
                            decreasing = T)
injuries_by_type <- sort(tapply(storm_data$INJURIES, storm_data$EVTYPE, sum),
                         decreasing = T)

fatal_cases_by_type[1:5]
injuries_by_type[1:5]

par(mfrow = c(2, 1))
barplot(fatal_cases_by_type[1:5], col = rgb(1, 0.8, 0.8), 
        ylab = "Number of fatal cases", cex.names = 0.7, space = 0,
        ylim = c(0, 6000))
barplot(injuries_by_type[1:5], col = rgb(0.8, 0.8, 1), 
        ylab = "Number of injuries", cex.names = 0.7, space = 0,
        ylim = c(0, 95000))
```
  
As we can see, tornado with 5633 fatalities and 91346 injuries
has significantly stronger harmful effects than any other type of event