---
title: "Storm Data Analysis"
author: "Levshinovskiy Mikhail"
date: "1/25/2021"
output:
  pdf_document: default
  html_document: default
---
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Synopsis

The analysis on the storm event data shows that the most harmful effect with
respect to population health have excessive heat with 1797 victims and tornado
with more than 20 000 injured people. The greatest economic damage are caused by
flood and drought, all together more that 150 billions dollar. The original dataset 
starts from 1950, but results were got based on data not earlier than 1996, because
earlier records were not fulfilled and results can't be relevant.


## Data Processing

Read dataset from archive, check if dimensions of data frame are correct. 
(Expected value: 902297 x 37)
```{r read_storm_data, cache=TRUE}
storm_data <- read.csv("./repdata_data_StormData.csv.bz2")
print(dim(storm_data))
names(storm_data)
```
Dataset description says that before *Jan. 1996* only tornado *EVTYPE* were recorded.
To compare damage and harmful effects of different events we need to subset data frame
by that date.
```{r subset_dataset_by_date}
storm_data$BGN_DATE <- as.Date(storm_data$BGN_DATE, "%m/%d/%Y %H:%M:%S")
storm_data <- storm_data[storm_data$BGN_DATE >= '1996-01-01',]
dim(storm_data)
```

Dataset description says that there are 48 types of event. Downloaded dataset has much more
```{r event_types_demo}
length(levels(storm_data$EVTYPE))
```


Some values in *EVTYPE* are different only because they are written with upper or
lower cases.  For example:
```{r upper_lower_example}
grep("[Cc][Oo][Aa][Ss][Tt][Aa][Ll] [Ss][Tt][Oo][Rr][Mm]", levels(storm_data$EVTYPE), value = T)
```
To reduce number of events that are different only because of case, change all of them to
lower case.
```{r to_lower}
storm_data$EVTYPE <- as.factor(tolower(storm_data$EVTYPE))
length(levels(storm_data$EVTYPE))
```

Finally we will try to check approximate matching of given *EVTYPES* with types from
description
```{r matching_EVTYPE_levels, cache=TRUE}
dict <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood",
          "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke",
          "Drought", "Dust Devil", "Dust Storm", "Excessive Heat",
          "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze",
          "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain",
          "Heavy Snow", "High Surf", "High Wind", "Hurricane (Typhoon)",
          "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", "Lightning",
          "Marine Hail", "Marine High Wind", "Marine Strong Wind",
          "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet",
          "Storm Surge/Tide", "Strong WInd", "Thunderstorm Wind", "Tornado",
          "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash",
          "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
dict <- tolower(dict)

library(stringdist)
get_stereotype <- function(ev) {
  ster <- amatch(ev, dict, maxDist = 4)
  if (!any(is.na(ster)))
    return(dict[ster])
  else
    return(NA)
}

storm_data$EVTYPE <- as.factor(sapply(storm_data$EVTYPE, get_stereotype))
storm_data <- storm_data[!is.na(storm_data$EVTYPE),]
levels(storm_data$EVTYPE)
```
## Results
  
### Question 1
  *Quesion* : Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
  
To answer this question we will sum number of injures and fatal cases for every event
type in dataset. Below numbers and *barplot* for 5 the most harmful events are presented.

```{r count_most_harmful_type}
fatal_cases_by_type <- sort(tapply(storm_data$FATALITIES, storm_data$EVTYPE, sum),
                            decreasing = T)
injuries_by_type <- sort(tapply(storm_data$INJURIES, storm_data$EVTYPE, sum),
                         decreasing = T)

fatal_cases_by_type[1:5]
injuries_by_type[1:5]

par(mfrow = c(2, 1))

barplot(fatal_cases_by_type[1:5], col = rgb(1, 0.8, 0.8), 
        ylab = "Number of fatal cases", cex.names = 0.8, space = 0,
        ylim = c(0, 1800))
barplot(injuries_by_type[1:5], col = rgb(0.8, 0.8, 1), 
        ylab = "Number of injuries", cex.names = 0.8, space = 0,
        ylim = c(0, 21000))
```



As we can see, excessive heat with 1797 fatalities and tornado with 91346 injuries
have significantly stronger harmful effects than any other type of event

### Question 2
  *Question* : Across the United States, which types of events have the greatest economic consequences?
  
Economic damage is presented with property damage and crop
damage. There are two fields for each type: *PRPDMG* and *PROPDMGEXP* for property damage
and *CROPDMG* and *CROPDMGEXP* for crop damage. To get damage in dollars we need to decode
values in these fields.
First function decode character from *PROPDMGEXP* and *CROPDMGEXP* fields to power of exponentiation.
```{r decodeing_function}
exp_decodeing <- function(ch) {
   # h -> hundred(2), k -> thousand(3), m -> million(6), b -> billion(9)
  if (ch %in% c('h', 'H'))
    return(2)
  if (ch %in% c('k', 'K'))
    return(3)
  if (ch %in% c('m', 'M'))
    return(6)
  if (ch %in% c('b', 'B'))
    return(9)
  if (!is.na(as.numeric(ch)))
    return(as.numeric(ch))
  if (ch %in% c('', '-', '?', '+'))
    return(1)
}
```

Now we can calculate damages for every type of event and draw barplots for top 5 of them.
```{r calculate_damage}
crop_dmg_exp <- sapply(storm_data$CROPDMGEXP, exp_decodeing)
storm_data$CROPDMG_dollars <- storm_data$CROPDMG * (10 ** crop_dmg_exp)
prop_dmg_exp <- sapply(storm_data$PROPDMGEXP, exp_decodeing)
storm_data$PROPDMG_dollars <- storm_data$PROPDMG * (10 ** prop_dmg_exp)

crop_dmg_dollars <- sort(tapply(storm_data$CROPDMG_dollars, storm_data$EVTYPE, sum), 
                         decreasing = T)
prop_dmg_dollars <- sort(tapply(storm_data$PROPDMG_dollars, storm_data$EVTYPE, sum),
                         decreasing = T)

crop_dmg_dollars[1:5]
prop_dmg_dollars[1:5]

par(fin = c(7, 5))
par(mfrow = c(2, 1))

barplot(crop_dmg_dollars[1:5], col = rgb(1, 0.8, 0.8), 
        ylab = "Crop damage in dollars", cex.names = 0.8, space = 0,
        oma = c(0.5, 0.1, 0.1, 0.1), cex.axis = 0.7)

barplot(prop_dmg_dollars[1:5], col = rgb(0.8, 0.8, 1), 
        ylab = "Property Damage in dollars", cex.names = 0.8, space = 0,
        oma = c(0.5, 0.1, 0.1, 0.1), cex.axis = 0.7)

```
The data shows that total crop damage from drought is more than 13.3 billions dollars and
property damage from flood is more than 143 billion dollars. These two
types of events have the greatest economic consequences.